---
title: "Queensland Road Safety"
author: 
  - Jacqueline Gunawan
  - Yu-Lin Chung
  - Kai Aditya Sharma
  - Kai-Hong Su 
format: 
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup

# Load libraries here
library(ggplot2)
library(dplyr)
library(knitr)
```

# Executive Summary

Max 4 sentences

# Introduction

Max 10 sentences

# Methodology

Max 300 words

```{r}
#| label: fig-method-plot
#| fig-cap: "Caption for methodology figure."

# plot code 
```

```{r}
#| label: tbl-method-table
#| tbl-cap: "Caption for methodology table."

# table code
```

# Results

Max 200 words

```{r}
# read data
raw_data <- read.csv("data/cleaned_dataset.csv")

# transform data
data <- raw_data|>
  mutate(
    Crash_Severity = factor(
      Crash_Severity,
      levels = c(
        "Minor injury",
        "Medical treatment",
        "Hospitalisation",
        "Fatal"
      ),
      ordered = TRUE
    ),
    Crash_Type = factor(Crash_Type),
    Crash_Nature = factor(Crash_Nature),
    Crash_Speed_Limit = factor(Crash_Speed_Limit),
    Crash_Road_Surface_Condition = factor(Crash_Road_Surface_Condition),
    Crash_Atmospheric_Condition = factor(Crash_Atmospheric_Condition),
    Crash_Lighting_Condition = factor(Crash_Lighting_Condition),
    Crash_Road_Horiz_Align = factor(Crash_Road_Horiz_Align),
    Crash_Road_Vert_Align = factor(Crash_Road_Vert_Align),
    Crash_DCA_Group_Description = factor(Crash_DCA_Group_Description)
  )

```

## How are crash severity outcomes distributed across high-speed road crashes in Brisbane in 2024?
```{r}
#| label: fig-crash-severity
#| fig-cap: "Distribution of Crash Severity Outcomes"

# plot for "How are crash severity outcomes distributed across high-speed road crashes in Brisbane in 2024?"
data |>
  count(Crash_Severity) |>
  mutate(
    percent = n / sum(n),
    label = scales::percent(percent, accuracy = 0.1)
  ) |>
  ggplot(aes(x = Crash_Severity, y = n)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.1) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Distribution of Crash Severity Outcomes",
    x = "Crash severity",
    y = "Number of crashes"
  )
```

@fig-crash-severity shows the overall distribution of crash severity outcomes. Hospitalisation and medical treatment crashes accounted for the largest shares, while fatal crashes represented only a small proportion of the dataset.


## Which crash types or crash natures are more commonly associated with serious outcomes?
```{r}
#| label: serious rate
#| fig-cap: "Serious Outcome Rate by Top Crash Natures"

top_natures <- data |>
  count(Crash_Nature, sort = TRUE) |>
  slice_head(n = 6) |>
  pull(Crash_Nature)

data |>
  mutate(
    Serious_Outcome = if_else(
      Crash_Severity %in% c("Hospitalisation", "Fatal"),
      "Serious",
      "Non-serious"
    )
  ) |>
  filter(Crash_Nature %in% top_natures) |>
  group_by(Crash_Nature) |>
  summarise(
  total_crashes = n(),
  serious_crashes = sum(Serious_Outcome == "Serious"),
  serious_rate = serious_crashes / total_crashes,
  .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(Crash_Nature, serious_rate), y = serious_rate)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(serious_rate, accuracy = 0.1)),
    hjust = -0.1
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Serious Outcome Rate by Top Crash Natures",
    x = "Crash nature",
    y = "Serious outcome rate"
  )
```
To examine which crash natures were more strongly associated with serious outcomes, the six most common crash natures were selected for comparison. Serious outcomes were defined as crashes resulting in hospitalisation or fatality.

# Discussion, Conclusion and Recommendations

# References
