---
title: "Crash Type and Severity on High-Speed Roads in Brisbane"
author:
  - "Jacqueline Gunawan"
  - "Yu-Lin Chung"
  - "Aditya Sharma"
  - "Kai-Hong Su"

format: revealjs
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup

library(ggplot2)
library(dplyr)
library(knitr)
library(tidyverse)
library(scales)
```

## Problem Introduction

-   Road traffic crashes cause approximately 1.19 million deaths globally each year [@who2023factsheet]
-   Speed is a key risk factor, with every 1% increase in mean speed producing a 4% increase in fatal crash risk [@who2023factsheet]
-   In 2024, Queensland recorded 303 road fatalities, the worst toll since 2009 [@qldgov2025]
-   High-speed roads are disproportionately associated with severe crash outcomes

## Dataset Description

- **Source:** Queensland Government Open Data Portal — *Crash data from Queensland roads*
- **Scope:** Brisbane police region, 2024 only
- **Filter:** Roads with speed limits of 80–110 km/h
- **Records:** 402 individual crash events, each with a unique reference number
- **20 variables** covering time, location, crash type, and environmental conditions
- **Outcome:** Crash Severity — Fatal, Hospitalisation, Medical treatment, or Minor injury
- **Quality:** No missing values, no duplicates, consistent formatting across all columns


## Methods

- We filtered the full Queensland dataset to Brisbane 2024, high-speed roads (80–110 km/h)
- Then, we checked data quality: zero missing values, no duplicates, all values within expected ranges
- No data was changed, removed, or filled in — dataset was already clean and complete
- We grouped severity into **Serious** (Fatal + Hospitalisation) and **Non-serious** (Medical treatment + Minor injury)
- Used bar charts and percentage comparisons to explore:
  - Overall crash severity distribution
  - Serious outcome rates by crash nature
  - Serious outcome rates by lighting and road surface conditions
  
## Results

### How are crash severity outcomes distributed across high-speed road crashes in Brisbane in 2024?

```{r}
raw_data <- read.csv("data/cleaned_dataset.csv")

data <- raw_data|>
  mutate(
    Crash_Severity = factor(
      Crash_Severity,
      levels = c(
        "Minor injury",
        "Medical treatment",
        "Hospitalisation",
        "Fatal"
      ),
      ordered = TRUE
    ),
    Crash_Type = factor(Crash_Type),
    Crash_Nature = factor(Crash_Nature),
    Crash_Speed_Limit = factor(Crash_Speed_Limit),
    Crash_Road_Surface_Condition = factor(Crash_Road_Surface_Condition),
    Crash_Atmospheric_Condition = factor(Crash_Atmospheric_Condition),
    Crash_Lighting_Condition = factor(Crash_Lighting_Condition),
    Crash_Road_Horiz_Align = factor(Crash_Road_Horiz_Align),
    Crash_Road_Vert_Align = factor(Crash_Road_Vert_Align),
    Crash_DCA_Group_Description = factor(Crash_DCA_Group_Description)
  )

```

```{r}
#| fig-cap: "Distribution of Crash Severity Outcomes for High-Speed Road Crashes"

# plot for "How are crash severity outcomes distributed across high-speed road crashes in Brisbane in 2024?"
data |>
  count(Crash_Severity) |>
  mutate(
    percent = n / sum(n),
    label = scales::percent(percent, accuracy = 0.1)
  ) |>
  ggplot(aes(x = Crash_Severity, y = n)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.1) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Distribution of Crash Severity Outcomes for High-Speed Road Crashes",
    x = "Crash severity",
    y = "Number of crashes"
  )
```

## Results

### How are crash severity outcomes distributed across high-speed road crashes in Brisbane in 2024?

- Hospitalisation was the most common outcome (41.3%).
- Medical treatment crashes were also frequent (39.1%).
- Fatal crashes were rare in the dataset (1.0%).
- Most high-speed road crashes involved some level of injury.

## Results

### Which crash types or crash natures are more commonly associated with serious outcomes?

```{r}
#| fig-cap: "Serious Outcome (hospitalisation or fatality) Rate by Top Crash Natures"

top_natures <- data |>
  count(Crash_Nature, sort = TRUE) |>
  slice_head(n = 6) |>
  pull(Crash_Nature)

data |>
  mutate(
    Serious_Outcome = if_else(
      Crash_Severity %in% c("Hospitalisation", "Fatal"),
      "Serious",
      "Non-serious"
    )
  ) |>
  filter(Crash_Nature %in% top_natures) |>
  group_by(Crash_Nature) |>
  summarise(
  total_crashes = n(),
  serious_crashes = sum(Serious_Outcome == "Serious"),
  serious_rate = serious_crashes / total_crashes,
  .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(Crash_Nature, serious_rate), y = serious_rate)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(serious_rate, accuracy = 0.1)),
    hjust = -0.1
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Serious Outcome Rate by Top Crash Natures",
    x = "Crash nature",
    y = "Serious outcome rate"
  )
```

## Results

### Which crash types or crash natures are more commonly associated with serious outcomes?

- Head-on crashes had the highest serious outcome rate (85.7%).
- Overturned and angle crashes also showed relatively high serious outcome rates.
- Rear-end crashes had the lowest serious outcome rate among the selected crash natures (35.9%).
- Crash frequency and crash severity do not necessarily follow the same pattern.

## Results

### Do lighting and road surface conditions appear to be related to crash severity?

```{r}
#| echo: false
#| fig-cap: "Serious Outcome Rate by Lighting and Road Surface Conditions"

condition_data <- data |>
  mutate(
    Serious_Outcome = if_else(
    Crash_Severity %in% c("Hospitalisation", "Fatal"),
    "Serious",
    "Non-serious"))

  condition_data <- condition_data |>
    select(
    Serious_Outcome,
    Crash_Lighting_Condition,
    Crash_Road_Surface_Condition) |>
      pivot_longer(
        cols = c(Crash_Lighting_Condition, Crash_Road_Surface_Condition),
        names_to = "Condition_Type",
        values_to = "Condition") 
        
  condition_data <- condition_data |>
    mutate(
      Condition_Type = case_when(
      Condition_Type == "Crash_Lighting_Condition" ~ "Lighting condition",
      Condition_Type == "Crash_Road_Surface_Condition" ~ "Road surface condition")) |>
        group_by(Condition_Type, Condition) |>
        summarise(
          total_crashes = n(),
          serious_crashes = sum(Serious_Outcome == "Serious"),
          serious_rate = serious_crashes / total_crashes,
          .groups = "drop") |>
            filter(total_crashes >= 5)

#plot
p <- condition_data |>
  ggplot(aes(x = reorder(Condition, serious_rate), y = serious_rate)) +
  geom_col(fill = "grey70") +
  geom_text(
  aes(label = scales::percent(serious_rate, accuracy = 0.1)),
  hjust = 1.1,
  color = "black")+
  coord_flip() +
  facet_wrap(~ Condition_Type, scales = "free_y") +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.2))) +
      labs(
        x = NULL,
        y = "Serious outcome rate"
        )

print(p)
```

## Results

### Do lighting and road surface conditions appear to be related to crash severity?

- Crashes occurring in darkness without lighting had the highest serious outcome rate at 52.6%.

- Crashes occurring on sealed wet roads showed a serious outcome rate than sealed dry roads of 46.2%. 

## Conclusions
- high-speed road crashes in Brisbane in 2024 were more commonly associated with injury outcomes than fatal outcomes.

- Crash nature was an important reason of crash severity. 
  - head-on crashes had the highest serious outcome rate

- Environmental conditions related to crash severity.
  - dark road without lighting had the highest serious outcome rate.
  - sealed wet roads showed a higher serious outcome rate than sealed dry roads. 

## Recommendations
- Road safety strategies should not only focus on preventing deaths , but also aim to reduce serious non-fatal injuries.
- To prevent serious crashes, agency can improve lighting in poorly lit areas, maintain road drainage systems, use wet-road warning signs and apply speed management strategies in high-risk locations.

## Reference