---
title: "Crash Type and Severity on High-Speed Roads in Brisbane"
author:
  - "Jacqueline Gunawan"
  - "Chung Yu Lin"
  - "Kai Aditya Sharma"
  - "Kai Hong Su"
format: 
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup

# Load libraries here
library(ggplot2)
library(dplyr)
library(knitr)
```


# Executive Summary

This report analyses 402 crashes recorded on high-speed roads (80 to 110 km/h) in the Brisbane police region in 2024, examining how crash type, crash nature, and environmental conditions relate to crash severity outcomes.
Head-on crashes and overturned vehicles were associated with the highest rates of serious outcomes, while crashes occurring in unlit darkness and on wet sealed roads also showed elevated serious outcome rates compared to other conditions.
These findings suggest that crash characteristics and environmental conditions play a meaningful role in determining crash severity on Brisbane's high-speed road network.
Road safety agencies should consider targeted interventions including improved lighting in high-risk locations, better road surface maintenance, and speed management strategies to reduce the frequency and severity of crashes on high-speed roads.



# Introduction

Road traffic crashes remain a critical public health challenge globally, with approximately 1.19 million people dying each year as a result of road traffic injuries [@who2023factsheet].
Road traffic injuries are also the leading cause of death for children and young adults aged 5 to 29 years, underscoring the urgency of targeted prevention efforts [@who2023factsheet].
Speed is a particularly significant risk factor, with every 1% increase in mean speed producing a 4% increase in fatal crash risk and a 3% increase in serious crash risk [@who2023factsheet].
High-speed roads are therefore disproportionately associated with severe crash outcomes, making them a priority area for road safety analysis.

Brisbane, as Queensland's most populous city, presents a complex road safety environment shaped by diverse road conditions and high-speed arterial networks operating at 80 to 110 km/h.
In 2024, Queensland recorded 303 fatalities on its roads, the worst toll since 2009, with speeding identified as a key contributing factor [@qldgov2025].
Prior research has shown that crash nature, lighting conditions, and road surface conditions are among the factors most associated with crash severity [@tamakloe2022].
This report analyses a dataset of 402 crashes recorded in the Brisbane police region in 2024, filtered to include only crashes on roads with speed limits of 80 to 110 km/h [@qlddata2024].
Key variables include crash severity, crash type, crash nature, road surface condition, and lighting condition.

The central research question guiding this analysis is: *which crash types and crash natures are more commonly associated with serious outcomes on high-speed roads in Brisbane?*


# Methodology

Max 300 words

```{r}
#| label: fig-method-plot
#| fig-cap: "Caption for methodology figure."

# plot code 
```

```{r}
#| label: tbl-method-table
#| tbl-cap: "Caption for methodology table."

# table code
```


# Results

Max 200 words

```{r}
#| label: fig-results-plot
#| fig-cap: "Caption for results figure."

# plot code
```


# Discussion, Conclusion and Recommendations

# References
