---
title: "Queensland Road Safety"
author: "Jacqueline Gunawan, Yu-Lin Chung, Kai Aditya Sharma, Kai Hong Su"
format: 
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup

# Load libraries here
library(ggplot2)
library(dplyr)
library(knitr)
```


# Executive Summary

Max 4 sentences


# Introduction

Max 10 sentences


# Methodology

Max 300 words

```{r}
#| label: fig-method-plot
#| fig-cap: "Caption for methodology figure."

# plot code 
```

```{r}
#| label: tbl-method-table
#| tbl-cap: "Caption for methodology table."

# table code
```


# Results

Max 200 words
```{r}
raw_data <- read.csv("data/cleaned_dataset.csv")
str(data)


data <- raw_data|>
  mutate(
    Crash_Severity = factor(
      Crash_Severity,
      levels = c(
        "Minor injury",
        "Medical treatment",
        "Hospitalisation",
        "Fatal"
      ),
      ordered = TRUE
    ),
    Crash_Type = factor(Crash_Type),
    Crash_Nature = factor(Crash_Nature),
    Crash_Speed_Limit = factor(Crash_Speed_Limit),
    Crash_Road_Surface_Condition = factor(Crash_Road_Surface_Condition),
    Crash_Atmospheric_Condition = factor(Crash_Atmospheric_Condition),
    Crash_Lighting_Condition = factor(Crash_Lighting_Condition),
    Crash_Road_Horiz_Align = factor(Crash_Road_Horiz_Align),
    Crash_Road_Vert_Align = factor(Crash_Road_Vert_Align),
    Crash_DCA_Group_Description = factor(Crash_DCA_Group_Description)
  )

```

# How are crash severity outcomes distributed across high-speed road crashes in Brisbane in 2024?
# Which crash types or crash natures are more commonly associated with serious outcomes?

# Do lighting and road surface conditions appear to be related to crash severity?
```{r}
#| fig-pos: "H"
#| label: fig-surface-light
#| fig-cap: "Serious Outcome Rate by Lighting and Road Surface Conditions"

library(tidyverse)
condition_data <- data |>
  mutate(
    Serious_Outcome = if_else(
    Crash_Severity %in% c("Hospitalisation", "Fatal"),
    "Serious",
    "Non-serious"))

  condition_data <- condition_data |>
    select(
    Serious_Outcome,
    Crash_Lighting_Condition,
    Crash_Road_Surface_Condition) |>
      pivot_longer(
        cols = c(Crash_Lighting_Condition, Crash_Road_Surface_Condition),
        names_to = "Condition_Type",
        values_to = "Condition") 
        
  condition_data <- condition_data |>
    mutate(
      Condition_Type = case_when(
      Condition_Type == "Crash_Lighting_Condition" ~ "Lighting condition",
      Condition_Type == "Crash_Road_Surface_Condition" ~ "Road surface condition")) |>
        group_by(Condition_Type, Condition) |>
        summarise(
          total_crashes = n(),
          serious_crashes = sum(Serious_Outcome == "Serious"),
          serious_rate = serious_crashes / total_crashes,
          .groups = "drop") |>
            filter(total_crashes >= 5)

#plot
condition_data |>
  ggplot(aes(x = reorder(Condition, serious_rate), y = serious_rate)) +
  geom_col(fill = "grey70") +
  geom_text(
  aes(label = scales::percent(serious_rate, accuracy = 0.1)),
  hjust = 1.1,
  color = "black")+
  coord_flip() +
  facet_wrap(~ Condition_Type, scales = "free_y") +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.2))) +
      labs(
        x = NULL,
        y = "Serious outcome rate"
        )
```



# Discussion, Conclusion and Recommendations

# References
