---
title: "Crash Type and Severity on High-Speed Roads in Brisbane"
author:
  - "Jacqueline Gunawan"
  - "Chung Yu Lin"
  - "Aditya Sharma"
  - "Kai Hong Su"

format: 
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup

# Loading our required libraries
library(ggplot2)
library(dplyr)
library(knitr)
library(readr)

crashes <- read_csv("data/cleaned_dataset.csv")
```


# Executive Summary

This report analyses 402 crashes recorded on high-speed roads (80 to 110 km/h) in the Brisbane police region in 2024, examining how crash type, crash nature, and environmental conditions relate to crash severity outcomes.
Head-on crashes and overturned vehicles were associated with the highest rates of serious outcomes, while crashes occurring in unlit darkness and on wet sealed roads also showed elevated serious outcome rates compared to other conditions.
These findings suggest that crash characteristics and environmental conditions play a meaningful role in determining crash severity on Brisbane's high-speed road network.
Road safety agencies should consider targeted interventions including improved lighting in high-risk locations, better road surface maintenance, and speed management strategies to reduce the frequency and severity of crashes on high-speed roads.


# Introduction

Road traffic crashes remain a critical public health challenge globally, with approximately 1.19 million people dying each year as a result of road traffic injuries.
Road traffic injuries are also the leading cause of death for children and young adults aged 5 to 29 years, underscoring the urgency of targeted prevention efforts.
Speed is a particularly significant risk factor, with every 1% increase in mean speed producing a 4% increase in fatal crash risk and a 3% increase in serious crash risk.
High-speed roads are therefore disproportionately associated with severe crash outcomes, making them a priority area for road safety analysis.

Brisbane, as Queensland's most populous city, presents a complex road safety environment shaped by diverse road conditions and high-speed arterial networks operating at 80 to 110 km/h.
In 2024, Queensland recorded 303 fatalities on its roads, the worst toll since 2009, with speeding identified as a key contributing factor.
Prior research has shown that crash nature, lighting conditions, and road surface conditions are among the factors most associated with crash severity.
This report analyses a dataset of 402 crashes recorded in the Brisbane police region in 2024, filtered to include only crashes on roads with speed limits of 80 to 110 km/h.
Key variables include crash severity, crash type, crash nature, road surface condition, and lighting condition.

The central research question guiding this analysis is: *which crash types and crash natures are more commonly associated with serious outcomes on high-speed roads in Brisbane?*



# Methodology

## Data Source

Our analysis is based on the *Crash data from Queensland roads* dataset,
which is publicly available through the Queensland Government open data
portal. The full dataset covers crash records across all of
Queensland over multiple years. To keep the population size manageable and
relevant to the scope of our analysis, we selected only records from the
Brisbane police region in 2024, along with crashes on roads with speed
limits between 80 and 110 km/h. This gave us a working dataset of 402
crash records. @fig-methodology-plot shows where these crashes are spread across
Brisbane.

## Variable Selection

The dataset contains 20 variables covering different aspects of each crash
(@tbl-methodology-table). These include details about when the crash happened
(month, day, hour), where it happened (latitude, longitude, postcode), how
it happened (crash nature, crash type, DCA group), and the conditions at
the time (speed limit, road surface, weather, lighting, road alignment).
The main outcome we focused on is *Crash_Severity*, which records whether
a crash resulted in a fatal, hospitalisation, medical treatment, or minor
injury outcome.

## Data Quality Checks

Before starting our analysis, we checked the filtered dataset for common
issues. We found no missing values in any of the 20 columns, no duplicate
records, and all text categories were consistent in spelling and
formatting. Numeric values such as crash hour (0 to 23), postcodes (4000
to 4520), and coordinates all fell within expected ranges for Brisbane.
Every crash reference number was unique, confirming that each row
represents a distinct crash event.

## Data Cleaning

After filtering, the *Crash_Year* and *Loc_Police_Region* columns held the
same value in every row (2024 and Brisbane) since we had already narrowed
the data to that scope. We kept these columns in the dataset for reference
but did not use them in the analysis. We did not change, remove, or fill
in any values during this process, as the data was already complete and
consistent.

## Analytical Approach

We used simple charts and summaries to explore how crash severity relates
to crash type, crash nature, and environmental conditions across
Brisbane's high-speed roads.

```{r}
#| label: tbl-methodology-table
#| tbl-cap: "Variables in the cleaned dataset, grouped by their role in the analysis."

var_dict <- data.frame(
  Role = c("Identifier", "Outcome", "Temporal",
           "Spatial", "Crash descriptors",
           "Environmental conditions", "Exposure"),
  Variables = c("Crash_Ref_Number",
                "Crash_Severity",
                "Crash_Month, Crash_Day_Of_Week, Crash_Hour",
                "Crash_Latitude, Crash_Longitude, Loc_Post_Code",
                "Crash_Nature, Crash_Type, Crash_DCA_Group_Description",
                "Crash_Speed_Limit, Road_Surface, Atmospheric, Lighting, Road_Horiz_Align, Road_Vert_Align",
                "Count_Unit_Car")
)

kable(var_dict, booktabs = TRUE)
```
\
```{r}
#| label: fig-methodology-plot
#| fig-cap: "Spatial distribution of the 402 crash records across the Brisbane police region in 2024, coloured by crash severity."
#| fig-width: 6
#| fig-height: 4

ggplot(crashes, aes(x = Crash_Longitude, y = Crash_Latitude,
                    colour = Crash_Severity)) +
  geom_point(alpha = 0.7, size = 1.6) +
  labs(x = "Longitude", y = "Latitude", colour = "Severity") +
  theme_minimal() +
  theme(legend.position = "bottom")
```


# Results

Max 200 words

```{r}
#| label: fig-results-plot
#| fig-cap: "Caption for results figure."

# plot code
```


# Discussion, Conclusion and Recommendations

# References
