---
title: "Queensland Road Safety"
author: "Jacqueline Gunawan", 
        "Chung Yu Lin", 
        "Kai Aditya Sharma", 
        "Kai Hong Su"
format: 
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup

# Load libraries here
library(ggplot2)
library(dplyr)
library(knitr)
```


# Executive Summary

Max 4 sentences


# Introduction

Max 10 sentences


# Methodology

Max 300 words

```{r}
#| label: fig-method-plot
#| fig-cap: "Caption for methodology figure."

# plot code 
```

```{r}
#| label: tbl-method-table
#| tbl-cap: "Caption for methodology table."

# table code
```


# Results

Max 200 words
```{r}
raw_data <- read.csv("data/cleaned_dataset.csv")
str(data)


data <- raw_data|>
  mutate(
    Crash_Severity = factor(
      Crash_Severity,
      levels = c(
        "Minor injury",
        "Medical treatment",
        "Hospitalisation",
        "Fatal"
      ),
      ordered = TRUE
    ),
    Crash_Type = factor(Crash_Type),
    Crash_Nature = factor(Crash_Nature),
    Crash_Speed_Limit = factor(Crash_Speed_Limit),
    Crash_Road_Surface_Condition = factor(Crash_Road_Surface_Condition),
    Crash_Atmospheric_Condition = factor(Crash_Atmospheric_Condition),
    Crash_Lighting_Condition = factor(Crash_Lighting_Condition),
    Crash_Road_Horiz_Align = factor(Crash_Road_Horiz_Align),
    Crash_Road_Vert_Align = factor(Crash_Road_Vert_Align),
    Crash_DCA_Group_Description = factor(Crash_DCA_Group_Description)
  )

```

# How are crash severity outcomes distributed across high-speed road crashes in Brisbane in 2024?
# Which crash types or crash natures are more commonly associated with serious outcomes?
# Do lighting and road surface conditions appear to be related to crash severity?
```{r}
#| label: fig-results-plot
#| fig-cap: "Caption for results figure."

# plot
data |>
  autoplot()
```


# Discussion, Conclusion and Recommendations

# References
